---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fastqq

<!-- badges: start -->
<!-- badges: end -->

`fastqq` is intended for creating quantile-quantile plots.
Initially intended for large-scale genome-wide association studies (GWAS). 
This is to investigate whether we have associations
that could be stronger than what one would expect by pure randomness.
In these cases, the user often plots tens to hundreds of millions of points.
Creating scatter plots with so many points is usually not efficient since the 
graphics devices store all the data, such that the visualization can be 
rescaled or plotted in a vector graphics format (where again all the data is stored).

A better and faster approach in these cases is to note that many of the points
are so close to each other that there is no value in including them in the
plot. QQ style plots are usually a monotonically increasing sequence of points,
so we can easily employ fast filtering to remove redundant points, that would
otherwise not be visible in the final plot anyways.

We provide faster alternatives to `qqman::qq`, `stats::qqplot` and 
`stats::qqnorm`. We also provide the function `fastqq::drop_dense` such that
the user can extract the data to plot with `ggplot`.

See examples below on how this can be achieved.

**Note** that this package is inspired by the `qqman` package, which has now
[been archived](https://github.com/stephenturner/qqman). The interface to the 
`qq` function should be very similar, and `fastqq::qq` should ideally be a 
drop-in replacement for `qqman::qq`. I created this package since it could take
30-60 minutes to render a single plot with `qqman::qq`.

## Installation

You can install the released version of `fastqq` from 
[CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("fastqq")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("gumeo/fastqq")
```
## Example

The following is an example from very simple simulated data:

```{r example}
library(fastqq)
set.seed(42)
p_simulated <- runif(10000)
# Classic way to do this with qqman
qqman::qq(p_simulated)

# Alternative
fastqq::qq(p_simulated)
```

We can compare the timings of creating the plots, with `qqman`.

```{r timing_fastqq}
set.seed(555)
N_test <- c(1e3,1e4,1e5,1e6)
time_method <- function(pkg_name, method){
  suppressPackageStartupMessages(library(pkg_name, 
                                         character.only=TRUE, quietly = TRUE))
  for(N in N_test){
    p_vec <- runif(n = N)
    print(paste0("Timing ", pkg_name, "::", method," with ", 
                 N, " points"))
    tictoc::tic()
    pdf(file = NULL) # Prevent the plots from appearing
    do.call(method, list(pvector=p_vec))
    dev.off()
    tictoc::toc()  
  }
}
N_test <- c(1e3,1e4,1e5,1e6)
time_method('fastqq','qq')
N_test <- c(1e3,1e4,1e5,1e6)
time_method('qqman','qq')
```

So we can expect around 8-10X speedup, and for 100 million points (order of magnitude for modern GWAS), `fastqq::qq` takes 30 seconds on the same hardware as for the timings above.

# Note on other efforts

After I created this, I have found several sources, that aim at something
similar, usually also a **manhattan** plot (I am probably also missing other packages):

- [`fastman`](https://github.com/roman-tremmel/ggfastman) package. Uses `scattermore`, so the plotting is very fast. This package is not currently (31/07/2021) on CRAN.
- [`ramwas`](https://github.com/andreyshabalin/ramwas) package. Has not been maintained in 2 years and is on bioconductor. This package is aimed for **Fast Methylome-Wide Association Study Pipeline for Enrichment Platforms** and the `ramwas::qqPlotFast` function is just a minor part of the package. It 

Here are also some projects on CRAN, where the plotting is similar to `qqman::qq`, and is not improved for speed.

- [`gwaRs`](https://github.com/LindoNkambule/gwaRs) package. Focuses on using `ggplot`.
- [`manhplot`](https://github.com/cgrace1978/manhplot) package. This is a pretty ambitious project, with a good [publication](https://bmcbioinformatics.biomedcentral.com/track/pdf/10.1186/s12859-019-3201-y.pdf). Again, the main focus is on the manhattan plot.
- [`qqman`](https://github.com/stephenturner/qqman) package. Probably one of the main inspiration for most of the other packages.
- [`CMplot`](https://github.com/YinLiLin/CMplot) package. Great circular manhattan plot.
